---
title: "ERG_data_analysis"
author: "Anna K"
date: "2026"
output: html_document
---

```{r variables, include=FALSE}

#Please set your pathnames here. Detailed instructions can be found in the quickstart guide.

input_folder = "FILL IN THE INPUT FOLDER PATH HERE"
input_metadata = "FILL IN THE METADATA PATH HERE"
output_folder = "FILL IN THE OUTPUT FOLDER PATH HERE"

#Please set your parameters here.

time_label = "Age (weeks)" #Define the time labels of your groups here. For example: "weeks since treatment", or "age".
grouping_var = "gender" #Please set the grouping variable of your choice. This is case sensitive! Check the quickstart guide for more info.

#Do you want to plot overviews of the applied filters? It is recommended to check these plots at least once, but it does take some time.

plot_all = TRUE

#Do you want to plot all waveforms? This will take more time, but allows you to identify potential outliers better. If so, set this variable to "TRUE"

identify_outliers = FALSE

#Please set this parameter to "TRUE" if you want to turn NA values into zeros (indicating an absence of a- or b-wave). Otherwise, they are kept as NA values, and they are ignored for statistical tests.

NA_zero = FALSE

#Change filters
#If you are working with different animal models/strains and want to tweak the filters, you can set them here. If the filters already work well, we recommend to not touch this setting.

lowpass = 30
highpass = 235

#Time settings
#Our standard ffERG programs make 512 measurements over 350 milliseconds. If you have other settings, change them here.

nr_timepoints = 512
timespan_perwave = 350

#If you have incompatible raw data and are uploading pre-processed data, set this to TRUE 

incompatible_data = FALSE
```

```{r setup, include=FALSE}

library(xlsx)

library(tidyverse)

library(signal)

library(flextable)

library(broom)

library(shiny)

library(DescTools)
```

```{r, include = FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
```

```{r load_data, results='hide'}
#load all the raw data

if(incompatible_data == FALSE){
files = list.files(path = input_folder, full.names = TRUE)

files_unique = files[!duplicated(basename(files))]

raw_data = files_unique %>% 
  map_dfr(~ read_csv(.x, show_col_types = FALSE))

}else if(incompatible_data == TRUE){
df = read.xlsx(input_folder, sheetIndex = 1)
}

patient_info = read.xlsx(input_metadata, sheetIndex = 1)
patient_info = patient_info[-c(8)]
```

```{r prep_data}
#prepare the data

if (incompatible_data == FALSE){

#this line of code works as follows: it scans the column 'clinic' until it finds the term 'Patient', then it pastes the line below (which is the patient number and is variable) in our new column "patient".
data = raw_data
data$patient  = ifelse(dplyr::lag(raw_data$Clinic,1,default = "NA") == "Patient",raw_data$Clinic,NA)

#do the same for the date
data$date  = ifelse(dplyr::lag(raw_data$...4,1,default = "NA") =="DateOfBirth,TestDate,PatId",raw_data$...4,NA)

#finding the range that is used by the system
range_nr  = ifelse(dplyr::lag(raw_data$Operator,1,default = "NA") =="Range(+/-[V])",raw_data$Operator,NA)
range_nr = range_nr[!is.na(range_nr)]
range_nr = unique(range_nr)

#fill all the line of codes in between different patient numbers with the last seen patient number.
data = data %>%
  fill(patient) 

#do the same for the date
data = data %>%
  fill(date) 

#find the stimfreq from the metadata rows
data = data %>%
  mutate(
    StimFreq = ifelse(
      Operator == range_nr,
      as.numeric(str_extract(...4, "[0-9.]+(?=,$)")),
      NA_real_
    )
  )


#in this chunk of code we do the following:
#1. in the column StimFreq, we find all numerical values that are back-to-back, until we reach an NA value
#2. We repeat the sequence of numbers 4 times in order
#3. Then we continue going down column StimFreq again until we find new numerical values
#4. Repeat until all numerical values were found
#5. Remove all NAs from the output

x = data$StimFreq

freqval = c()
i = 1
n = length(x)

while (i <= n) {
  
  # Step 1: get block of non-NA values
  if (!is.na(x[i])) {
    block_start = i
    
    while (i <= n && !is.na(x[i])) {
      i = i + 1
    }
    
    block = x[block_start:(i-1)]
    
    # Step 2: repeat block 4 times
    freqval = c(freqval, rep(block, 4))
    
  } else {
    # skip NA rows
    freqval = c(freqval, x[i])
    i = i + 1
  }
}
##

freqval = na.omit(freqval)


#change labels of columns
names(data)[names(data) == "Address"] = "Measurement_type"
names(data)[names(data) == "...4"] = "Volt"

#keep only rows that contain actual measurements (removes titles and summaries). 
#The "Operator !=" command is to remove the lines that give the estimation of the a- and b-waves from the machine
#And the voltage has to contain any number
numbers = c("0", "1", "2", "3", "4", "5", "6", "7", "8", "9")
data = data %>% dplyr::filter(grepl("-",Measurement_type) & Operator != range_nr & grepl(paste(numbers,collapse="|"), Volt))

data$StimFreq = freqval

#remove noise from the date 
data$date = gsub("(.*),.*", "\\1", as.character(data$date))
data$date = gsub(".*,","",as.character(data$date))

#merge with patient data
names(data)[names(data) == "patient"] = "identifier"
names(data)[names(data) == "date"] = "PatId"

data = merge(data,patient_info,by=c("identifier","PatId"))

#remove the pause info
data = data %>% dplyr::filter(!grepl("Pause",Measurement_type))

}else if (incompatible_data == TRUE){
  print("Pre-processed data was uploaded, moving to analysis.")
}

```


# Data plots

### Raw data

```{r create_df} 

#calculate the different timepoints in ms
time_jumps = timespan_perwave/nr_timepoints
time_points = seq(from = 0, to = timespan_perwave, by = time_jumps)
time_points = time_points[-c(1)] #0 is not included.

#make dataframe with timepoints and the volt measurements.

if (incompatible_data == FALSE){
#order the data so we don't mess up the graph
data = data[order(data$Operator),]

#combine all the rows that belong to 1 singular measurement into 1 row
df = data %>%
  group_by(identifier, Measurement_type, StimFreq, PatId) %>%
  summarise(Volt = str_c(Volt, collapse = ","))

} 

#make every separate voltage measurement a separate row. this is automatically in order.
df = df %>% 
  mutate(Volt = strsplit(as.character(Volt), ",")) %>%
  unnest(Volt)

#remove NAs
df$Volt = as.numeric(df$Volt)
df = na.omit(df)

#add timepoints
df$Time = rep_len(time_points, length.out = length(df$Volt))

#quick plot to test data quality
test_plot = ggplot(data=df, aes(x=Time, y=Volt, color=PatId)) +
  facet_grid(Measurement_type+StimFreq~identifier)+
  geom_line(show.legend = FALSE)
test_plot

```

### Zero-centered data

```{r zero_center} 
#filtering the data by zero-centering and applying two low-pass filters.

#calculate avg voltage of measurements before flash (<20ms)
subtract_data = df%>%
  group_by(identifier, Measurement_type, StimFreq, PatId)%>%
  dplyr::filter(Time <= 20)

#take the average of each group and make new column subtract_val with this value
subtract_value = subtract_data%>%
  group_by(identifier, Measurement_type, StimFreq, PatId)%>%
  mutate(subtract_val = mean(Volt, na.rm=TRUE))

#remove rows that specify things we dont need (we only want the patient, measurement, and date to be associated with a certain subtract value)
subtract_value = subset(subtract_value, select = -c(Volt, Time) )

#remove duplicate rows
subtract_value= unique(subtract_value)

#merge the regular data frame with the subtract value to get zero-centered (0c) data
df_0c = merge(df,subtract_value,by=c("identifier","Measurement_type", "StimFreq", "PatId"))

df_0c$Volt_0c = df_0c$Volt-df_0c$subtract_val

df_0c = subset(df_0c, select = -c(Volt) )

#simple plot of zero-centered data
if (plot_all == TRUE){
test_plot_0c = ggplot(data=df_0c, aes(x=Time, y=Volt_0c, color=PatId)) +
  facet_grid(Measurement_type+StimFreq~identifier)+
  geom_line(show.legend = FALSE)

test_plot_0c
}
```


```{r filter} 

#new df that omits the flicker data

df_oneflash = df_0c %>% dplyr::filter(StimFreq < 1 ) #stimfreq >1 means it is a flicker

#low-pass filter creatie en toepassing
#4th order, 30Hz (1/30) en 235Hz (1/235)

bf1 = butter(4, 1/lowpass, type="low")

Volt_lpf= filtfilt(bf1, df_oneflash$Volt_0c) #lpf = low-pass filtered

bf2 = butter(4, 1/highpass, type="high")

Volt_hpf= filtfilt(bf2, df_oneflash$Volt_0c)

#add new data to our df
df_oneflash$Volt_lpf= Volt_lpf
df_oneflash$Volt_hpf= Volt_hpf

```

### Filtered data (flash)

1. zero-centered, removed flicker.
2. low-pass filtered (for b-wave)
3. high-pass filtered (for a-wave)

```{r check_filter}

#quick plots to check the filtering
if (plot_all == TRUE){
  
plot_0c = ggplot(data=df_oneflash, aes(x=Time, y=Volt_0c, color=PatId)) +
  facet_grid(Measurement_type~identifier)+
  geom_line(show.legend = FALSE)

plot_0c

plot_lpf = ggplot(data=df_oneflash, aes(x=Time, y=Volt_lpf, color=PatId)) +
  facet_grid(Measurement_type~identifier)+
  geom_line(show.legend = FALSE)

plot_lpf

plot_hpf = ggplot(data=df_oneflash, aes(x=Time, y=Volt_hpf, color=PatId)) +
  facet_grid(Measurement_type~identifier)+
  geom_line(show.legend = FALSE)

plot_hpf

}else{
  print("Overview plots were not created.")
}

```

### Plotting the data per animal to identify potential outliers
#This is not automatically plotted. If you want to plot all waveforms, set "identify_outliers" to TRUE.


```{r identify_outliers}

if (identify_outliers == TRUE){

all_traces = df_0c

all_traces =  merge(all_traces,patient_info,by=c("identifier","PatId"))

all_traces$Measurement_info = paste(all_traces$Measurement_type, all_traces$StimFreq)

all_traces %>%
  group_split(identifier) %>%
  map(~ {
    group_name = unique(.x$identifier)
    
    ggplot(.x, aes(x = Time, y = Volt_0c, color = as.factor(timespan))) +
      geom_line() +
      facet_wrap(~ Measurement_info) +
      labs(title = group_name, legend = time_label)
  })
}else if (identify_outliers == FALSE){
  print("Individual waveforms were not printed")
}
```




```{r find_ab-waves}
#finding a and b waves in the filtered data

#find all peaks
local_maximas = df_oneflash[which(diff(sign(diff(df_oneflash$Volt_lpf)))==-2)+1,]

local_maximas = subset(local_maximas, select = c(identifier, Measurement_type, StimFreq, PatId, Volt_lpf, Time) )
names(local_maximas)[names(local_maximas) == "Volt_lpf"] = "maximas"

#keep only peaks that occured after the stimulus
local_maximas = local_maximas %>%
  dplyr::filter(Time>20)

#find the highest peak (b-wave)
b_waves = local_maximas %>%
  group_by(identifier, Measurement_type, StimFreq, PatId)%>%
  top_n(1, maximas) #highest peak

#make a new df for the times of the b-waves
bwave_time = b_waves

names(bwave_time)[names(bwave_time) == "Time"] = "time_bwave"

bwave_time = subset(bwave_time, select = -c(maximas) )

#find valleys
local_minimas = df_oneflash[which(diff(sign(diff(df_oneflash$Volt_hpf)))==+2)-1,]

local_minimas = subset(local_minimas, select = c(identifier, Measurement_type, StimFreq, PatId, Volt_hpf, Time) )
names(local_minimas)[names(local_minimas) == "Volt_hpf"] = "minimas"

#add the b-wave time to the df, so we can compare the times (a-wave must be before the b-wave)
local_minimas = merge(local_minimas,bwave_time,by=c("identifier","Measurement_type", "StimFreq", "PatId"))

#make sure the a-wave is before the b-wave and after the stimulus
a_waves = local_minimas %>%
  dplyr::filter(Time>20&Time<time_bwave)

#find the lowest valley (a-wave)
a_waves = a_waves %>%
  group_by(identifier, Measurement_type, StimFreq, PatId)%>%
  top_n(1, -minimas) #deepest valleys

  


```


```{r data_waves, include=FALSE}
#getting data for statistics

#finding the true a-wave amplitude by interpolating the zero-centered data and looking up the timepoint we found on the filtered data.


list_awave = list()

for (i in 1:nrow(a_waves)){
  row = a_waves[i,]
  subset = dplyr::filter(df_0c, identifier == row$identifier & Measurement_type == row$Measurement_type & StimFreq == row$StimFreq & PatId == row$PatId)
  print(nrow(subset))
  awave_volt = approx(subset$Time, subset$Volt_0c, row$Time, method="linear")
  list_awave[[i]] = awave_volt$y
  print(ifelse(nrow(subset) == 512, "Done! Continuing...", "ERROR! Wrong number of measurements."))
}
print("Finished processing a-wave amplitudes")


#finding the true b-wave amplitude by interpolating the zero-centered data and looking up the timepoint we found on the filtered data

list_bwave = list()

for (i in 1:nrow(b_waves)){
  rowb = b_waves[i,]
  subsetb = dplyr::filter(df_0c, identifier == rowb$identifier & Measurement_type == rowb$Measurement_type & StimFreq == rowb$StimFreq & PatId == rowb$PatId)
  print(nrow(subsetb))
  bwave_volt = approx(subsetb$Time, subsetb$Volt_0c, rowb$Time, method="linear")
  list_bwave[[i]] = bwave_volt$y
  print(ifelse(nrow(subset) == 512, "Done! Continuing...", "ERROR! Wrong number of measurements."))
}
print("Finished processing b-wave amplitudes")

#optionally turn NA values to zero. You can set this at the start of the program.

if (NA_zero == TRUE){
  list_awave[is.na(list_awave)] = 0
  list_bwave[is.na(list_bwave)] = 0
  print("NA values were set to zero.")
} else {
  print("NA values were maintained.")
}

#calculate b-wave amplitude

list_awave = unlist(list_awave)
list_bwave = unlist(list_bwave)

#create the dataframe with our points of interest!

a_waves$awave_true = list_awave
b_waves$bwave_true = list_bwave

names(b_waves)[names(b_waves) == "Time"] = "time_bwave"
names(a_waves)[names(a_waves) == "Time"] = "time_awave"
a_waves = subset(a_waves, select = -c(time_bwave) )

waves = merge(a_waves, b_waves, by=c("identifier", "Measurement_type", "StimFreq", "PatId"), all=TRUE)

waves$NAto0 = waves$awave_true

waves["NAto0"][is.na(waves["NAto0"])] = 0

ERG_data = data.frame(identifier = waves$identifier, Eye=waves$Measurement_type, StimFreq= waves$StimFreq, PatId =waves$PatId, awave = -(waves$awave_true), bwave = waves$bwave_true - waves$NAto0, bwave_abs =  waves$bwave_true, awave_imp_time = waves$time_awave-20, bwave_imp_time = waves$time_bwave-20, awave_abs_time = waves$time_awave, bwave_abs_time = waves$time_bwave)

ERG_data[c('eye', 'type', 'stimulus')] = str_split_fixed(ERG_data$Eye, ' ', 3)

ERG_data = merge(ERG_data,patient_info,by=c("identifier","PatId"))

ERG_data = ERG_data %>% mutate (treatment = 
                   case_when(eye == "L-2" ~ treatment_left,
                             eye == "R-1" ~ treatment_right,
                             .default = NA))

ERG_data = subset(ERG_data, select = -c(treatment_left, treatment_right) )

ERG_data = ERG_data[c('identifier', 'gender', 'genotype', 'treatment', 'PatId', 'timespan', 'eye', 'type', 'stimulus', 'StimFreq', 'awave', 'bwave', 'bwave_abs', 'awave_imp_time', 'bwave_imp_time', 'awave_abs_time', 'bwave_abs_time')]


```

###Calculating and visualizing AUC


```{r AUC, include=FALSE}


list_AUC = list()

for (i in 1:nrow(waves)){
  rowAUC = waves[i,]
  subsetAUC = dplyr::filter(df_0c, identifier == rowAUC$identifier & Measurement_type == rowAUC$Measurement_type & StimFreq == rowAUC$StimFreq & PatId == rowAUC$PatId)
  print(nrow(subsetAUC))
  xAUC = subsetAUC$Time
  yAUC = abs(subsetAUC$Volt_0c)
  AUC = AUC(xAUC, yAUC)
  list_AUC[[i]] = AUC
  print(ifelse(nrow(subset) == 512, "Done! Continuing...", "ERROR! Wrong number of measurements."))
}

list_AUC = unlist(list_AUC)

ERG_data$AUC = list_AUC

ERG_data$timespan = as.factor(ERG_data$timespan)

auc_pic = ggplot(data=ERG_data, aes(x=timespan, y=AUC)) +
  facet_grid(stimulus+StimFreq~.data[[grouping_var]])+
  geom_boxplot(outlier.colour="red", outlier.shape = 8, outlier.size = 3)+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5, aes(color=identifier))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  labs(title="AUC after different stimuli",
        x ="Weeks since treatment", y = "b-wave (V)")

auc_pic
```




### Filtered data (flicker) 

1. zero-centered data
2. low-pass filtered
3. high-pass filtered

```{r flicker_filter} 

#now selecting flicker data

df_flicker = df_0c %>% dplyr::filter(StimFreq > 1 ) #stimfreq >1 means it is a flicker


plot_flicker = ggplot(data=df_flicker, aes(x=Time, y=Volt_0c, color=PatId)) +
  facet_grid(Measurement_type~identifier)+
  geom_line(show.legend = FALSE)

plot_flicker

#low-pass filter creatie en toepassing
#4th order, 30Hz (1/30) en 235Hz (1/235)

bf1 = butter(4, 1/30, type="low")

flicker_lpf= filtfilt(bf1, df_flicker$Volt_0c) #lpf = low-pass filtered

bf2 = butter(4, 1/235, type="high")

flicker_hpf= filtfilt(bf2, flicker_lpf)

#add new data to our df
df_flicker$flicker_lpf= flicker_lpf
df_flicker$flicker_hpf= flicker_hpf
```


```{r check_flicker_filter} 

flicker_plot_hpf = ggplot(data=df_flicker, aes(x=Time, y=flicker_hpf, color=PatId)) +
  facet_grid(Measurement_type~identifier)+
  geom_line(show.legend = FALSE)

flicker_plot_hpf

```


```{r flicker_peaks}
#finding peaks and freq of flicker data

#find peaks
flicker_maximas = df_flicker[which(diff(sign(diff(df_flicker$flicker_hpf)))==-2)+1,]

flicker_maximas = subset(flicker_maximas, select = c(identifier, Measurement_type, StimFreq, PatId, flicker_hpf, Time) )
names(flicker_maximas)[names(flicker_maximas) == "flicker_hpf"] = "maximas"

flicker_maximas = flicker_maximas %>%
  dplyr::filter(Time>20)

#find the top 3 peaks -- this should already be achieved by filtering time>20, but is just a failsafe in case of weird ERG data.
flicker_peaks = flicker_maximas %>%
  group_by(identifier, Measurement_type, StimFreq, PatId)%>%
  top_n(3, maximas) #highest peaks

#find valleys
flicker_minimas = df_flicker[which(diff(sign(diff(df_flicker$flicker_hpf)))==+2)-1,]

flicker_minimas = subset(flicker_minimas, select = c(identifier, Measurement_type, StimFreq, PatId, flicker_hpf, Time) )
names(flicker_minimas)[names(flicker_minimas) == "flicker_hpf"] = "minimas"

#find the best valleys
flicker_valleys = flicker_minimas %>%
  group_by(identifier, Measurement_type, StimFreq, PatId)%>%
  top_n(3, -minimas) #deepest valleys


```

```{r stats_flicker, include=FALSE}
#getting data for statistics

#finding the frequency of the flicker response waves

flicker_stats = flicker_peaks %>%
  group_by(identifier, StimFreq, PatId, Measurement_type) %>%
  summarise_at(vars(Time), list(min=min, max=max))

flicker_stats$freq = 1000/((flicker_stats$max-flicker_stats$min)/2)

#calculating the average amplitude of the waves

#first, find the true values of peaks and valleys
list_flickerpeak = list()

for (i in 1:nrow(flicker_peaks)){
  row = flicker_peaks[i,]
  subset = dplyr::filter(df_flicker, identifier == row$identifier & StimFreq == row$StimFreq & Measurement_type == row$Measurement_type & PatId == row$PatId)
  flickerpeak_volt = approx(subset$Time, subset$Volt_0c, row$Time, method="linear")
  list_flickerpeak[[i]] = flickerpeak_volt$y
  print(ifelse(nrow(subset) == 512, "Done! Continuing...", "ERROR! Wrong number of measurements."))
}
print("Finished processing flicker waveform peaks.")

list_flickerpeak = unlist(list_flickerpeak)

flicker_peaks$maximas_0c = list_flickerpeak

flicker_peak2 = flicker_peaks %>%
  group_by(identifier, Measurement_type, StimFreq, PatId) %>%
  dplyr::filter(n() == 1L | row_number() == 2L) %>%
  ungroup()

list_flickerval = list()

for (i in 1:nrow(flicker_valleys)){
  row = flicker_valleys[i,]
  subset = dplyr::filter(df_flicker, identifier == row$identifier & StimFreq == row$StimFreq & Measurement_type == row$Measurement_type & PatId == row$PatId)
  flickerval_volt = approx(subset$Time, subset$Volt_0c, row$Time, method="linear")
  list_flickerval[[i]] = flickerval_volt$y
  print(ifelse(nrow(subset) == 512, "Done! Continuing...", "ERROR! Wrong number of measurements."))
}
print("Finished processing flicker waveform valleys.")

list_flickerval = unlist(list_flickerval)

flicker_valleys$minimas_0c = list_flickerval

flicker_val2 = flicker_valleys %>%
  group_by(identifier, Measurement_type, StimFreq, PatId) %>%
  dplyr::filter(n() == 1L | row_number() == 2L) %>%
  ungroup()

#now add the data to a frame and sum peaks and valleys together
flicker_statspeak = flicker_peaks %>%
  group_by(identifier, StimFreq, PatId, Measurement_type) %>%
  summarise_at(vars(maximas_0c), list(sum_peaks = sum))

flicker_statsval = flicker_valleys %>%
  group_by(identifier, StimFreq, PatId, Measurement_type) %>%
  summarise_at(vars(minimas_0c), list(sum_vals = sum))

flicker_statsamp = merge(flicker_statsval,flicker_statspeak,by=c("identifier","Measurement_type","StimFreq", "PatId"))

#find avg amplitude by subtracting valleys from peaks and dividing by 3 (nr of peaks/valleys)
flicker_statsamp$avg_amp = (flicker_statsamp$sum_peaks - flicker_statsamp$sum_vals)/3

flicker_statsamp = subset(flicker_statsamp, select = c(identifier, Measurement_type, StimFreq, PatId, avg_amp) )

#find amp of second peak

flicker_2 = merge(flicker_peak2, flicker_val2, by=c("identifier","Measurement_type", "StimFreq","PatId"))

flicker_2$peak2_amp = flicker_2$maximas_0c - flicker_2$minimas_0c

flicker_2 = subset(flicker_2, select = c(identifier, Measurement_type, StimFreq, PatId, peak2_amp) )

#adding this data to the stats dataset
flicker_stats = merge(flicker_stats,flicker_statsamp,by=c("identifier","Measurement_type", "StimFreq","PatId"))
flicker_stats = merge(flicker_stats,flicker_2,by=c("identifier","Measurement_type", "StimFreq","PatId"))

#renaming for clarity
names(flicker_stats)[names(flicker_stats) == "min"] = "first_peak_time"

#removing uninteresting data
flicker_stats = subset(flicker_stats, select = -c(max) )

flicker_stats[c('eye', 'type', 'stimulus')] = str_split_fixed(flicker_stats$Measurement_type, ' ', 3)

flicker_stats = flicker_stats[c('identifier', 'PatId', 'eye', 'type', 'StimFreq','stimulus', 'first_peak_time', 'avg_amp', 'freq', 'peak2_amp')]

#adding patient data
flicker_stats = merge(flicker_stats,patient_info,by=c("identifier","PatId"))

flicker_stats = flicker_stats %>% mutate (treatment = 
                   case_when(eye == "L-2" ~ treatment_left,
                             eye == "R-1" ~ treatment_right,
                             .default = NA))

flicker_stats = subset(flicker_stats, select = -c(treatment_left, treatment_right) )

flicker_stats = flicker_stats[c('identifier', 'gender', 'genotype', 'treatment', 'StimFreq', 'PatId', 'timespan', 'eye', 'type', 'stimulus', 'first_peak_time', 'avg_amp', 'freq', 'peak2_amp')]

write.xlsx(ERG_data, file = file.path(output_folder, "testtest.xlsx"), sheetName="flash", row.names=FALSE)
write.xlsx(flicker_stats, file = file.path(output_folder, "testtest.xlsx"), sheetName="flicker", append=TRUE, row.names=FALSE)

```


### ANOVA for b-wave amplitude

#### Graphical view of data (b-wave amplitude)
*Shown are the average b-wave amplitudes per group with SE errorbars. The numbers at the top describe the light intensity.*

```{r bwave_stats, results='hide'}
ERG_data$.data[[grouping_var]]=as.factor(ERG_data$.data[[grouping_var]])
ERG_data$timespan=as.factor(ERG_data$timespan)

ERG_data_scotopic = ERG_data %>% dplyr::filter(type == "scotopic")
ERG_data_photopic = ERG_data %>% dplyr::filter(type == "photopic")


data_summary = function(data, varname, groupnames){
  require(plyr)
  
  summary_func = function(x, col){
    vals = x[[col]]
    n = sum(!is.na(vals))
    sd_val = sd(vals, na.rm = TRUE)
    
    c(
      mean = mean(vals, na.rm = TRUE),
      n = n,
      sd = sd_val,
      SE = sd_val / sqrt(n)
    )
  }
  
  data_sum = ddply(data, groupnames, .fun = summary_func, varname)
  data_sum = rename(data_sum, c("mean" = varname))
  return(data_sum)
}
  

bwave_data = data_summary(ERG_data, varname="bwave", 
                    groupnames=c("stimulus", "timespan", "type","genotype", "gender", "treatment"))


  
bwave_stimulus = ggplot(data=bwave_data, aes(x=timespan, y=bwave*1000000, group=.data[[grouping_var]], color=.data[[grouping_var]])) +
  facet_wrap(.~stimulus+type, scales="free_y")+
  geom_line(aes(color=.data[[grouping_var]]))+
  geom_point()+
  geom_errorbar(aes(ymin=bwave*1000000-SE*1000000, ymax=bwave*1000000+SE*1000000), width=.1)+
  labs(title="b-wave amplitude after different stimuli",
        x =time_label, y = "b-wave (μV)")+
  theme_linedraw()

bwave_stimulus 

#ANOVA formula
anova_formula <- as.formula(
  paste("bwave ~", grouping_var, "* timespan")
)

# run ANOVA for each stimulus
anova_results = ERG_data %>%
  group_by(stimulus, type) %>%
  nest() %>%
  mutate(
    model = map(data, ~ aov(anova_formula, data = .x)),
    anova_table = map(model, broom::tidy)
  )

anova_summary = anova_results %>%
  select(stimulus, type, anova_table) %>%
  unnest(anova_table) %>%
  mutate(
    p.value.display = paste0(
      format.pval(
        p.value,
        digits = 2,
        eps = 0.001,
        nsmall = 3
      ),
      
      case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.01  ~ "**",
        p.value < 0.05  ~ "*",
        TRUE ~ ""
      )
    )
  )

#run Tukey
tukey_results <- anova_results %>%
  mutate(
    tukey = map(model, ~ TukeyHSD(.x)),
    
    tukey_table = map(tukey, ~ {
      broom::tidy(.x) %>%
        dplyr::filter(adj.p.value < 0.05) %>%
        mutate(
          adj.p.value = paste0(
            format.pval(
              adj.p.value,
              digits = 2,
              eps = 0.001,
              nsmall = 3
            ),
            
            case_when(
              as.numeric(adj.p.value) < 0.001 ~ "***",
              as.numeric(adj.p.value) < 0.01  ~ "**",
              as.numeric(adj.p.value) < 0.05  ~ "*",
              TRUE ~ ""
            )
          )
        )
    })
  )

tukey_results = tukey_results %>%
  mutate(
    flextable = map(tukey_table, flextable)
  )

all_tukey = tukey_results %>%
  select(stimulus, type, tukey_table) %>%
  unnest(tukey_table)

write.xlsx(as.data.frame(all_tukey),file = file.path(output_folder, "testtest.xlsx"), sheetName="bwave_pval", append=TRUE, row.names=FALSE)

```

***

#### Results of ANOVA (b-wave amplitude)

model information:
* assumes equal groups (all combinations of timespan x grouping_var should have equal participants)
* assumes an interaction between timespan x grouping_var
* assumes normality and homoscedasticity. 

The ANOVA is run separately for each stimulus.

```{r anova_bwave}
anova_summary
```

***

#### Comparisons of all groups (b-wave amplitude)
##### *only significant results are shown*

```{r anova_tuk_bwave}

all_tukey
```

***

#### Plots to check data quality (b-wave amplitude)

##### *Residuals vs fitted should be a horizontal line with data points distributed more or less evenly around it. Visible numbers are outliers. Consider removing these.*

##### *Q-Q plot should show all data points approximately in a straight line at a 45 degree angle. Large skews mean the data are not normally distributed. Consider a non-parametric approach.*

*note: if you have many (near-)blind rats and just a few healthy rats, expect the QQ plot to be heavy-tailed and there to be a few outliers.*

for more information on Q-Q plots and how to interpret them, check this [link](https://www.ucd.ie/ecomodel/Resources/QQplots_WebVersion.html)

```{r bwave_normality}

anova_results %>%
  mutate(
    plot_title = paste("Stimulus:", stimulus,
                       "| Type:", type)
  ) %>%
  pwalk(function(model, plot_title, ...) {

    par(mfrow = c(1, 2))  # side-by-side plots

    # Residuals vs fitted
    plot(
      model,
      which = 1,
      main = paste("Residuals vs Fitted\n", plot_title)
    )

    # QQ plot
    plot(
      model,
      which = 2,
      main = paste("QQ Plot\n", plot_title)
    )
  })

```

### ANOVA for a-wave amplitude

#### Graphical view of data (a-wave amplitude)
*Shown are the average b-wave amplitudes per group with SE errorbars. The numbers at the top describe the light intensity.*

```{r awave_stats, results='hide'}

awave_data = data_summary(ERG_data, varname="awave", 
                    groupnames=c("stimulus", "timespan", "type","genotype", "gender", "treatment"))


awave_stimulus = ggplot(data=awave_data, aes(x=timespan, y=awave*1000000, group=.data[[grouping_var]], color=.data[[grouping_var]])) +
  facet_wrap(.~stimulus+type, scales="free_y")+
  geom_line(aes(color=.data[[grouping_var]]))+
  geom_point()+
  geom_errorbar(aes(ymin=awave*1000000-SE*1000000, ymax=awave*1000000+SE*1000000), width=.1)+
  labs(title="a-wave amplitude after different stimuli",
        x =time_label, y = "a-wave (μV)")+
    theme_linedraw()

awave_stimulus


#ANOVA formula
anova_formula_awave <- as.formula(
  paste("awave ~", grouping_var, "* timespan")
)

# run ANOVA for each stimulus
anova_results_awave = ERG_data %>%
  group_by(stimulus, type) %>%
  nest() %>%
  mutate(
    model = map(data, ~ aov(anova_formula_awave, data = .x)),
    anova_table = map(model, broom::tidy)
  )

anova_summary_awave = anova_results_awave %>%
  select(stimulus, type, anova_table) %>%
  unnest(anova_table) %>%
  mutate(
    p.value.display = paste0(
      format.pval(
        p.value,
        digits = 2,
        eps = 0.001,
        nsmall = 3
      ),
      
      case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.01  ~ "**",
        p.value < 0.05  ~ "*",
        TRUE ~ ""
      )
    )
  )


#run Tukey
tukey_results_awave = anova_results_awave %>%
  mutate(
    tukey = map(model, ~ TukeyHSD(.x)),
    
    tukey_table = map(tukey, ~ {
      broom::tidy(.x) %>%
        dplyr::filter(adj.p.value < 0.05) %>%
        mutate(
          adj.p.value = paste0(
            format.pval(
              adj.p.value,
              digits = 2,
              eps = 0.001,
              nsmall = 3
            ),
            
            case_when(
              as.numeric(adj.p.value) < 0.001 ~ "***",
              as.numeric(adj.p.value) < 0.01  ~ "**",
              as.numeric(adj.p.value) < 0.05  ~ "*",
              TRUE ~ ""
            )
          )
        )
    })
  )

tukey_results_awave = tukey_results_awave %>%
  mutate(
    flextable = map(tukey_table, flextable)
  )

all_tukey_awave = tukey_results_awave %>%
  select(stimulus, type, tukey_table) %>%
  unnest(tukey_table)



write.xlsx(as.data.frame(all_tukey_awave),file = file.path(output_folder, "testtest.xlsx"), sheetName="awave_pval", append=TRUE, row.names=FALSE)
```

***

#### Results of ANOVA (a-wave amplitude)

model information:
* assumes equal groups (all combinations of timespan x grouping_var should have equal participants)
* assumes an interaction between timespan x grouping_var
* assumes normality and homoscedasticity. 

```{r anova_awave}
anova_summary_awave
```

***

#### Comparisons of all groups (a-wave amplitude)
##### *only significant results are shown*

```{r anova_tuk_awave}

all_tukey_awave
```

***

#### Plots to check data quality (a-wave amplitude)

##### *Residuals vs fitted should be a horizontal line with data points distributed more or less evenly around it. Visible numbers are outliers. Consider removing these.*

##### *Q-Q plot should show all data points approximately in a straight line at a 45 degree angle. Large skews mean the data are not normally distributed. Consider a non-parametric approach.*

*note: if you have many (blind) knockout rats and just a few WT rats, expect the QQ plot to be heavy-tailed and there to be a few outliers.*

for more information on Q-Q plots and how to interpret them, check this [link](https://www.ucd.ie/ecomodel/Resources/QQplots_WebVersion.html)

```{r awave_normality}

anova_results_awave %>%
  mutate(
    plot_title = paste("Stimulus:", stimulus,
                       "| Type:", type)
  ) %>%
  pwalk(function(model, plot_title, ...) {

    par(mfrow = c(1, 2))  # side-by-side plots

    # Residuals vs fitted
    plot(
      model,
      which = 1,
      main = paste("Residuals vs Fitted\n", plot_title)
    )

    # QQ plot
    plot(
      model,
      which = 2,
      main = paste("QQ Plot\n", plot_title)
    )
  })

```


### ANOVA for flicker peak2 amplitude

#### Graphical view of data (flicker peak2 amplitude)


```{r flicker_stats, results='hide'}
flicker_stats$.data[[grouping_var]]=as.factor(flicker_stats$.data[[grouping_var]])
flicker_stats$timespan=as.factor(flicker_stats$timespan)

flicker_data = data_summary(flicker_stats, varname="avg_amp", 
                    groupnames= c("stimulus", "timespan", "StimFreq","type", "genotype", "gender", "treatment"))

flicker_stimulus = ggplot(data=flicker_data, aes(x=timespan, y=avg_amp*1000000, group=.data[[grouping_var]], color=.data[[grouping_var]])) +
  geom_line(aes(color=.data[[grouping_var]]))+
  facet_wrap(.~stimulus+StimFreq+type, labeller = label_both)+
  geom_point()+
  geom_errorbar(aes(ymin=avg_amp*1000000-SE*1000000, ymax=avg_amp*1000000+SE*1000000), width=.1)+
  labs(title="average amplitude after flicker stimulus",
        x =time_label, y = "Average amplitude (μV)")+
  theme_linedraw()

flicker_stimulus

flicker_data$.data[[grouping_var]]=as.factor(flicker_data$.data[[grouping_var]])
flicker_data$timespan=as.factor(flicker_data$timespan)

#ANOVA formula
anova_formula_flicker <- as.formula(
  paste("avg_amp ~", grouping_var, "* timespan")
)

# run ANOVA for each stimulus
anova_results_flicker = flicker_stats %>%
  group_by(stimulus, type) %>%
  nest() %>%
  mutate(
    model = map(data, ~ aov(anova_formula_flicker, data = .x)),
    anova_table = map(model, broom::tidy)
  )

anova_summary_flicker = anova_results_flicker %>%
  select(stimulus, type, anova_table) %>%
  unnest(anova_table) %>%
  mutate(
    p.value.display = paste0(
      format.pval(
        p.value,
        digits = 2,
        eps = 0.001,
        nsmall = 3
      ),
      
      case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.01  ~ "**",
        p.value < 0.05  ~ "*",
        TRUE ~ ""
      )
    )
  )


#run Tukey
tukey_results_flicker = anova_results_flicker %>%
  mutate(
    tukey = map(model, ~ TukeyHSD(.x)),
    
    tukey_table = map(tukey, ~ {
      broom::tidy(.x) %>%
        dplyr::filter(adj.p.value < 0.05) %>%
        mutate(
          adj.p.value = paste0(
            format.pval(
              adj.p.value,
              digits = 2,
              eps = 0.001,
              nsmall = 3
            ),
            
            case_when(
              as.numeric(adj.p.value) < 0.001 ~ "***",
              as.numeric(adj.p.value) < 0.01  ~ "**",
              as.numeric(adj.p.value) < 0.05  ~ "*",
              TRUE ~ ""
            )
          )
        )
    })
  )

tukey_results_flicker = tukey_results_flicker %>%
  mutate(
    flextable = map(tukey_table, flextable)
  )

all_tukey_flicker = tukey_results_flicker %>%
  select(stimulus, type, tukey_table) %>%
  unnest(tukey_table)


#write.xlsx(as.data.frame(all_tukey_flicker),file = file.path(output_folder, "testtest.xlsx"), sheetName="flicker_pval", append=TRUE, row.names=FALSE)

```

***

#### Results of ANOVA (flicker peak2 amplitude)

```{r anova_flicker}
anova_summary_flicker
```

***

#### Comparisons of all groups (flicker amplitude)
##### *only significant results are shown*

```{r anova_tuk_flicker}

all_tukey_flicker
```

***

#### Plots to check data quality (flicker amplitude)

##### *Residuals vs fitted should be a horizontal line with data points distributed more or less evenly around it. Visible numbers are outliers. Consider removing these.*

##### *Q-Q plot should show all data points approximately in a straight line at a 45 degree angle. Large skews mean the data are not normally distributed. Consider a non-parametric approach.*

*note: if you have many (blind) knockout rats and just a few WT rats, expect the QQ plot to be heavy-tailed and there to be a few outliers.*

for more information on Q-Q plots and how to interpret them, check this [link](https://www.ucd.ie/ecomodel/Resources/QQplots_WebVersion.html)

```{r flicker_normality}
anova_results_flicker %>%
  mutate(
    plot_title = paste("Stimulus:", stimulus,
                       "| Type:", type)
  ) %>%
  pwalk(function(model, plot_title, ...) {

    par(mfrow = c(1, 2))  # side-by-side plots

    # Residuals vs fitted
    plot(
      model,
      which = 1,
      main = paste("Residuals vs Fitted\n", plot_title)
    )

    # QQ plot
    plot(
      model,
      which = 2,
      main = paste("QQ Plot\n", plot_title)
    )
  })

```














